```{r include = FALSE}
child_setup <- file.path("../basic/00setup.Rmd")
child_close <- file.path("../basic/01close.Rmd")
```

```{r child=child_setup}
```

## Introduction

Now we will use the clipping functions. There are two ways of
clipping data in ICESat2VegR:

  1. Clipping raw hdf5 data from ATL03 and ATL08 files
  2. Clipping extracted attributes resulting from extraction functions:
    - `ATL03_seg_attributes_dt`
    - `ATL03_photon_attributes_dt`
    - `ATL08_seg_attributes_dt`
    - `ATL03_ATL08_photons_attributes_dt_join`

The second method is preffered as it is faster and more efficient
because it does not require reading the hdf5 and will clip only
the subset of extracted attributes, while hdf5 has a lot of attributes
that might not be needed.

There are multiple variants using either the bounding box or the 
geomtry to clip the data suffixed with _clipBox or _clipGeometry:

  1. `ATL03_h5_clipBox`
  2. `ATL03_h5_clipGeometry`
  3. `ATL03_seg_attributes_dt_clipBox`
  4. `ATL03_seg_attributes_dt_clipGeometry`
  5. `ATL03_photon_attributes_dt_clipBox`
  6. `ATL03_photon_attributes_dt_clipGeometry`
  7. `ATL08_h5_clipBox`
  8. `ATL08_h5_clipGeometry`
  9. `ATL08_seg_attributes_dt_clipBox`
  10. `ATL08_seg_attributes_dt_clipGeometry`
  11. `ATL03_ATL08_photons_attributes_dt_join_clipBox`
  12. `ATL03_ATL08_photons_attributes_dt_join_clipGeometry`

In the following two sections there are two small examples
on how to clip the raw HDF5 and the extracted attributes.

## Clipping raw hdf5 data from ATL08

```{r, message = FALSE, fig.width = 5, out.width = '260', fig.height = 8, fig.align = "center"}
# Define bbox
clip_region <- terra::ext(-83.2, -83.14, 32.12, 32.18)

# Define hdf5 output file
output <- tempfile(fileext = ".h5")

# Clip the data for only the first atl08 file
atl08_clipped <- ATL08_h5_clipBox(atl08_h5[[1]], output, bbox = clip_region)

atl08_seg_dt <- ATL08_seg_attributes_dt(atl08_h5[[1]], attributes = c("h_canopy"))
atl08_seg_dt_clip <- ATL08_seg_attributes_dt(atl08_clipped, attributes = c("h_canopy"))

# Display location of clipped data
plot(
  atl08_seg_dt$longitude,
  atl08_seg_dt$latitude,
  pch = 20,
  xlab = "longitude",
  ylab = "latitude"
)
points(atl08_seg_dt_clip$longitude, atl08_seg_dt_clip$latitude, col = "purple", pch = 3, cex = 1)
rect(clip_region$xmin, clip_region$ymin, clip_region$xmax, clip_region$ymax, lty = 2)

legend(col = c("black", "purple", "black"), legend = c("Original", "Clipped", "AOI"), pch = c(20, 3, NA), lty = c(NA, NA, 2), cex = 1, x = "topright")
```

## Clipping extracted attributes from ATL08 segments data
```{r, fig.width = 5, out.width = 260, fig.height = 9, fig.align = "center"}
aoi <- file.path(outdir, "example_aoi.gpkg")
aoi_vect <- terra::vect(aoi)

# Extract the h_canopy attribute from the first ATL08 file
atl08_seg_dt <- lapply(atl08_h5, ATL08_seg_attributes_dt, attributes = c("h_canopy"))
atl08_seg_dt <- rbindlist2(atl08_seg_dt)

# Clip the data for only the first atl08 file
atl08_seg_dt_clip <- ATL08_seg_attributes_dt_clipGeometry(atl08_seg_dt, aoi_vect)

plot(
  atl08_seg_dt$longitude,
  atl08_seg_dt$latitude,
  pch = 20,
  xlab = "longitude",
  ylab = "latitude"
)
points(atl08_seg_dt_clip$longitude, atl08_seg_dt_clip$latitude, col = "purple", pch = 3, cex = 1)
aoi_geom <- terra::geom(aoi_vect)
polygon(aoi_geom[,"x"], aoi_geom[,"y"], lty = 2)
legend(col = c("black", "purple", "black"), legend = c("Original", "Clipped", "AOI"), pch = c(20, 3, NA), lty = c(NA, NA, 2), cex = 1, x = "topright")
```


```{r child=child_close}
```
