## Introduction

There are two different ways of working with the ICESat-2 data. Locally or using cloud computing. Common users should work locally, unless they are working within an AWS cloud computing within zone us-west-2.

```{r include = FALSE}
hide_close08 <- TRUE
if (!exists("vignettes_path")) {
  vignettes_path <- ".."
}

child_setup <- file.path("../basic/00setup.Rmd")
child_close <- file.path("../basic/01close.Rmd")
```

```{r child=child_setup}
```


## Search parameters

```{r bbox}
# Specifying bounding box coordinates
lower_left_lon <- -96.0
lower_left_lat <- 40.0
upper_right_lon <- -100
upper_right_lat <- 42.0


# Specifying the date range
daterange <- c("2021-10-02", "2021-10-03")
```

## Working locally

First we need to find the granules:

```{r local, cache=FALSE}
atl03_granules_local <- ATLAS_dataFinder(
  short_name = "ATL03",
  lower_left_lon,
  lower_left_lat,
  upper_right_lon,
  upper_right_lat,
  version = "006",
  daterange = daterange,
  persist = TRUE,
  cloud_computing = FALSE
)

head(atl03_granules_local)
```

Now we download the granules:

```{r eval = FALSE}
# Download all granules
ATLAS_dataDownload(atl03_granules_local, "/the/directory/to/save")
```

And then we can open and work with them

```{r eval = FALSE}
# Read the granules
atl03_h5 <- ATL03_read("/the/directory/to/save/name_of_granule.h5")

# List groups within atl03_h5
atl03_h5$ls()
```

```{r read, echo = FALSE}
# List groups within atl03_h5
atl03_h5[[1]]$ls()
```

## Working in the cloud
```{r cloud_find, cache=FALSE}
atl03_granules_cloud <- ATLAS_dataFinder(
  short_name = "ATL03",
  lower_left_lon,
  lower_left_lat,
  upper_right_lon,
  upper_right_lat,
  version = "006",
  daterange = daterange,
  persist = TRUE,
  cloud_computing = TRUE
)

head(atl03_granules_cloud)
```

In cloud computing you don't need to download data, instead you can 
read the data and start working with it.

```{r, cache=FALSE}
# Read the granule (the ATL03_read can only read one granule per read)
atl03_h5_cloud <- ATL03_read(atl03_granules_cloud[1])

# List groups within the h5 in cloud
atl03_h5_cloud$ls()
```

```{r, cache=FALSE}
# Which are strong beams
print(atl03_h5_cloud$strong_beams)
```

```{r, cache=FALSE}
# Orientation 0=forward, 1=backwards, 2=transition
print(atl03_h5_cloud[["orbit_info/sc_orient"]][])
```

```{r child=child_close}
```

```{r}
close(atl03_h5_cloud)
```
