```{r include = FALSE}
child_setup <- file.path("../basic/00setup.Rmd")
child_close <- file.path("../basic/01close.Rmd")
```

## Introduction

In this section, we will demonstrate how to use the
`ATL03_ATL08_segment_create` function from the `ICESat2VegR` package. 
This function is used to compute segment IDs for ICESat-2 `ATL03` and `ATL08`
data and create segments based on a specified segment length.

```{r child=child_setup}
```

```{r, eval = FALSE}
# Herein as we are working with list of h5 files we will need
# to loop over each file and extract the attributes and then
# concatenate them with rbindlist2
atl03_atl08_dts <- list()

for (ii in seq_along(atl03_h5)) {
  atl03_file <- atl03_h5[[ii]]
  atl08_file <- atl08_h5[[ii]]

  atl03_atl08_dts[[ii]] <- ATL03_ATL08_photons_attributes_dt_join(
    atl03_file,
    atl08_file
  )
}

atl03_atl08_dt <- rbindlist2(atl03_atl08_dts)

head(atl03_atl08_dt)
```


## Create Segments IDs

Now, we use the `ATL03_ATL08_segment_create` function to create segments with a specified segment length.


```{r}
atl03_atl08_photons_grouped_dt <- ATL03_ATL08_segment_create(atl03_atl08_dt,
  segment_length = 30,
  centroid = "mean",
  output = NA,
  overwrite = FALSE
)
```

## Compute Segment Statistics
```{r, results = 'hide'}
atl03_atl08_seg_dt <- ATL03_ATL08_compute_seg_attributes_dt_segStat(
  atl03_atl08_photons_grouped_dt,
  list(
    h_canopy_ge0 = quantile(ph_h, 0.98),
    h_canopy_gt0 = quantile(ph_h[ph_h > 0], 0.98),
    n_ground = sum(classed_pc_flag == 1),
    n_mid_canopy = sum(classed_pc_flag == 2),
    n_top_canopy = sum(classed_pc_flag == 3),
    n_canopy_total = sum(classed_pc_flag >= 2)
  ),
  ph_class = c(1, 2, 3)
)

head(atl03_atl08_seg_dt)
```

<div align="center" style="overflow-x: scroll;">

```{r echo = FALSE, results = 'asis'}
print(head(atl03_atl08_seg_dt))
```

</div>


## Convert to SpatVector

Now, we convert the data.table to a SpatVector object.

```{r}
atl03_atl08_vect <- to_vect(atl03_atl08_seg_dt[h_canopy_gt0 <= 31])
```

## Visualize the segments

Finally, we visualize the SpatVector interactively using mapview.


```{r, eval = c(1:3)}
centroid <- atl03_atl08_dt[, .(x = mean(lon_ph), y = mean(lat_ph))]

map_output <- mapview::mapview(
  atl03_atl08_vect,
  zcol = "h_canopy_gt0",
  col.regions = grDevices::hcl.colors(9, "RdYlGn"),
  alpha = 0,
  layer.name = "h_canopy",
  map.types = c("Esri.WorldImagery"),
  cex = 4
)@map %>%
  setView(lng = centroid$x, lat = centroid$y, zoom = 13)

map_output
```

<div align="center">

```{r segments_h_canopy, echo = FALSE, fig.cap = "Leaflet preview for segments h_canopy", out.height='400', out.width='500', fig.align = "center"}
if (requireNamespace("mapview")) {
  printMapView(map_output, "atl03_atl08_vect", output = TRUE, delay = 10)
}
```

</div>


```{r child=child_close}
```