---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
suppressWarnings(rm("first_setup_run"))
if (!exists("vignettes_path")) {
  vignettes_path <- "."
}
```

```{r installation, eval = FALSE}
# The r-universe version (recommended for the latest version)
install.packages("ICESat2VegR", , repos = c("https://caiohamamura.r-universe.dev", "https://cloud.r-project.org"))

# The CRAN version
install.packages("ICESat2VegR")
```

## Load the package

```{r eval = FALSE, echo = TRUE}
library(ICESat2VegR)
```


## Configuring the package

This package uses three Python packages through `reticulate`:

1. [earthaccess](https://github.com/nsidc/earthaccess): allows reading directly from the cloud
2. [h5py](https://github.com/h5py/h5py): for reading hdf5 content from the cloud
3. [earthengine-api](https://github.com/google/earthengine-api): integration with Google Earth Engine for sampling and extracting raster data and upscalling models.

For configuring the package you can use:

```{r configure, eval = FALSE}
ICESat2VegR_configure()
```

This will install miniconda if not available and the necessary packages.

### Notes

 - There are some issues regarding some Python packages not being compatible with the Python version. The above configure function will also try to update python version in that case. 
 - The configure function may warn you about the need to restart R after installing some packages,
 restart if needed.