---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{v01_Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (!exists("basic_path")) {
  basic_path <- "basic"
}
```

```{r installation, eval = FALSE}
# The CRAN version
install.packages("ICESat2VegR")

# Development version
remotes::install_github("https://github.com/carlos-alberto-silva/ICESat2VegR")
```

```{r child=file.path(basic_path,"01load.Rmd")}
```

## Load the package

```{r eval = FALSE, echo = TRUE}
library(ICESat2VegR)
```


## Configuring the package

This package uses three Python packages through `reticulate`:

1. [earthaccess](https://github.com/nsidc/earthaccess): allows reading directly from the cloud
2. [h5py](https://github.com/h5py/h5py): for reading hdf5 content from the cloud
3. [earthengine-api](https://github.com/google/earthengine-api): integration with Google Earth Engine for sampling and extracting raster data and upscalling models.

For configuring the package you can use:

```{r configure, eval = FALSE}
ICESat2VegR_configure()

```

This will install miniconda if not available and the necessary packages.

### Notes

 - There are some issues regarding some Python packages not being compatible with the Python version. The above configure function will also try to update python version in that case. 
 - The configure function will warn you about the need to restart R